Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

[EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks. This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

A lot of epidemiological research is interested in understanding intrinsic properties of the epidemic, not depending upon exactly when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$, a time-invariant value that is [defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population.] Interestingly, \citet{Gallagher2020} has shown that the estimation of $R_0$  [can be sensitive to time-dependent parameters] like estimation of [the beginning and end of an epidemic]. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproducible number. One common time-dependent tool is a series of time series line plots that track the proportion of the population in a given state (e.g. infected) at a given time. Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (s)usceptible, (i)nfectious and (r)ecovered. These simulations are actually generated under a discrete approximation of  \citet{Kermack1927}’s SIR model. [This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,]

@@

[From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.]

@@

Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare. Even though these two simulated epidemics appear different in these figures ([including having different infection peaks]), there is no real intuitive way to compare the epidemic’s $R_0$. In fact, both of these simulated epidemics started out the same population (1000 people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$). Even with simple generative models, this time-dependent visualization tools is not able to be used to compare a very simple but powerful numerical summary.

@@

The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state could be examined at a given time.  A more holistic approach to visualizing the overall path of our simulated epidemics would be to examine each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space. Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space. [For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.] In this both of Figure  \ref{fig:different-scales-tern}’s plots we an observe that the epidemics are on the same trajectory. In this simple generative example setting, [this indicates that \ben{the} two epidemics have the same $R_0$ value]. This can be proven mathematically.

@@

\begin{theorem}\label{thm:sir-scale}
 Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$. [Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.]
\end{theorem}

[The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model. In other words, the two epidemics form the same filamental trajectory.]

@@

[Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  better compare the overall structure of the epidemic and see how the population is directly impacted.] In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states. These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so. This approach is useful when the epidemic of interest has only gone [through a single realization of its outbreak (before the population of individuals become susceptible again)].

The first set of tools allows a practioner to define distances between epidemics whose time features don’t well align. For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points. [This representation induces a natural distance \ben{\sout{between this type of representation}} between epidemics, specifically:]

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
\]
where $\tilde{\psi}_i(s)$ the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics,] and practitioners can interpret the univariate epidemic's extremeness score relative to the extremeness scores of the simulations very easily.

When the goal is to predict the future impact / trajectory of a epidemic, time-invariant tools can help describe state-space regions in which we expect the true epidemic to traverse. In settings where the epidemic only goes a single outbreak, these regions can be very telling if simulation models well capture the epidemic’s structure. In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated epidemics.  These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances. Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.

>>>>

EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks. This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.  This intrinsic property\footnote{added to shorten this long sentence and connect that it is an intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}. \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number. One common time-dependent tool \shannon{used to better understand epidemics} is a series of time series line plots that track the proportion of the population in a given state (e.g. infected) at a given time. Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (S)usceptible, (I)nfectious and (R)ecovered. These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model. The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.

@@

Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare. Even though these two simulated epidemics appear different in these figures (including having different infection peaks), there is no real intuitive way to compare the epidemics' $R_0$ values. In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$). Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, $R_0$.

@@

The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.  A more holistic approach to visualizing the overall path of our simulated epidemics is to examine how each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space. Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space. For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot. In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space. In this simple generative example setting, being on the same trajectory indicates that \ben{the} two epidemics have the same $R_0$ value. This can be proven mathematically.

@@

\begin{theorem}\label{thm:sir-scale}
 Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e. $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$. Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$, then every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model. In other words, the two epidemics form the same filamental trajectory.

@@

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space (i.e. through a time-invariant lens) allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted. In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states. These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so. This approach is useful when the epidemic of interest has only gone through a single realization of its outbreak (before the population of individuals become susceptible again).

The first set of tools allows a practioner to define distances between epidemics whose time features do not align. For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points. This representation induces a natural distance between epidemics, specifically:

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
\]
where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$, where $|\psi_i|$ is the length of $\psi_i$. This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.

\sout{When the goal is to predict the future impact / trajectory of a epidemic,} In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse. In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure. In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated trajectories.  These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances. Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.
